gratia: An R package for exploring generalized additive models

### 2024-12-20 paper.Rmd
fix a couple of typos

---
title: 'gratia: An R package for exploring generalized additive models'
tags:
  - R
  - splines
  - generalized additive models
  - posterior simulation
authors:
  - name: Gavin L. Simpson
    orcid: 0000-0002-9084-8413
    affiliation: 1
affiliations:
 - name: Department of Animal and Veterinary Science, Aarhus University, Denmark
   index: 1
date: 26 June 2024
citation_author: Simpson
year: 2024
bibliography: paper.bib
output:
  rticles::joss_article:
    keep_md: true
csl: apa.csl
journal: JOSS
---

@@

@@

```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 5, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline. a) shows the individual basis functions (thin coloured lines; numbers indicate which basis function each line represents) multiplied by their respective model coefficients, as well as the data (black points) to which the GAM was fitted. The estimated smooth is shown as the thick grey line. b) shows the penalty matrix for the basis shown in a). Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0.", fig.pos = "t!"}
pkgs <- c("mgcv", "gratia", "ggplot2", "geomtextpath", "dplyr", "tibble", "patchwork")
res <- vapply(pkgs, library, logical(1), character.only = TRUE,
  logical.return = TRUE)
# simulate data
sim_d <- tibble(x = withr::with_seed(1234, runif(100))) |>
  mutate(
    f = gw_f2(x),
    y = withr::with_seed(1234, rnorm(100, mean = f, sd = 1))
  )
# fit model
sim_m <- gam(y ~ s(x, bs = "tp"), data = sim_d, method = "REML",
  drop.intercept = FALSE)
# basis functions for s(x)
bfs <- basis(sim_m, select = "s(x)")
# penalty S
f_x <- bfs |>
  group_by(x) |>
  summarise(.spline = sum(.value), .groups = "drop")
# recentre on the data
bfs <- bfs |>
  mutate(.value = .value + model_constant(sim_m))
f_x <- f_x |>
  mutate(.spline = .spline + model_constant(sim_m))
# plot basis functions and fitted spline
p_bfs <- bfs |>
  ggplot(aes(x = x, y = .value)) +
  geom_point(data = sim_d, aes(x = x, y = y)) +
  geom_line(data = f_x, aes(x = x, y = .spline),
    linewidth = 1.25, alpha = 0.5) +
  geom_labelline(
    aes(x = x, y = .value, colour = .bf, label = .bf),
    hjust = "auto",
    label.padding = unit(0.15, "lines"),
    size = 3
  ) +
  guides(colour = "none") +
  labs(x = "x", y = "y", title = "s(x)")
# plot the penalty
S <- penalty(sim_m, select = "s(x)")
p_S <- S |>
  draw()
p_bfs + p_S + plot_layout(ncol = 2, nrow = 1) +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

@@

*gratia* provides a range of functionality to make working with estimated GAMs easier for users. It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposing the powerful machinery of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects. As such, it provides a useful addition for anyone wanting to use GAMs in their data analyses without requiring them to be a GAM expert.

>>>>

---
title: 'gratia: An R package for exploring generalized additive models'
tags:
  - R
  - splines
  - generalized additive models
  - posterior simulation
authors:
  - name: Gavin L. Simpson
    orcid: 0000-0002-9084-8413
    affiliation: 1
affiliations:
 - name: Department of Animal and Veterinary Sciences, Aarhus University, Denmark
   index: 1
date: 26 June 2024
citation_author: Simpson
year: 2024
bibliography: paper.bib
output:
  rticles::joss_article:
    keep_md: true
csl: apa.csl
journal: JOSS
---

@@

```{r, setup, echo = FALSE, cache = FALSE, results = "hide"}
pkgs <- c("mgcv", "gratia", "ggplot2", "geomtextpath", "dplyr", "tibble", "patchwork")
res <- vapply(pkgs, library, logical(1), character.only = TRUE,
  logical.return = TRUE)
```

@@

```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 5, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline. a) shows the individual basis functions (thin coloured lines; numbers indicate which basis function each line represents) multiplied by their respective model coefficients, as well as the data (black points) to which the GAM was fitted. The estimated smooth is shown as the thick grey line. b) shows the penalty matrix for the basis shown in a). Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0.", fig.pos = "t!"}
# simulate data
sim_d <- tibble(x = withr::with_seed(1234, runif(100))) |>
  mutate(
    f = gw_f2(x),
    y = withr::with_seed(1234, rnorm(100, mean = f, sd = 1))
  )
# fit model
sim_m <- gam(y ~ s(x, bs = "tp"), data = sim_d, method = "REML",
  drop.intercept = FALSE)
# basis functions for s(x)
bfs <- basis(sim_m, select = "s(x)")
# penalty S
f_x <- bfs |>
  group_by(x) |>
  summarise(.spline = sum(.value), .groups = "drop")
# recentre on the data
bfs <- bfs |>
  mutate(.value = .value + model_constant(sim_m))
f_x <- f_x |>
  mutate(.spline = .spline + model_constant(sim_m))
# plot basis functions and fitted spline
p_bfs <- bfs |>
  ggplot(aes(x = x, y = .value)) +
  geom_point(data = sim_d, aes(x = x, y = y)) +
  geom_line(data = f_x, aes(x = x, y = .spline),
    linewidth = 1.25, alpha = 0.5) +
  geom_labelline(
    aes(x = x, y = .value, colour = .bf, label = .bf),
    hjust = "auto",
    label.padding = unit(0.15, "lines"),
    size = 3
  ) +
  guides(colour = "none") +
  labs(x = "x", y = "y", title = "s(x)")
# plot the penalty
S <- penalty(sim_m, select = "s(x)")
p_S <- S |>
  draw()
p_bfs + p_S + plot_layout(ncol = 2, nrow = 1) +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

@@

*gratia* provides a range of functionality to make working with estimated GAMs easier for users. It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposes the powerful machinery of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects. As such, it provides a useful addition for anyone wanting to use GAMs in their data analyses without requiring them to be a GAM expert.
